---
title: "Real-time analysis of the 2018 Ebola outbreak in the Democratic Republic of Congo"
output: github_document
---

14 June 2018

Christian L. Althaus (christian.althaus@alumni.ethz.ch), [Immuno-Epidemiology](http://www.immuno-epidemiology.ch/), Institute of Social and Preventive Medicine, University of Bern, Bern, Switzerland

## Introduction

```{r init, include=FALSE}
# init
rm(list = ls())

# Necessary libraries
library(deSolve)
library(chron)
library(bbmle)
library(mvtnorm)

# Initialize random number generator
set.seed(12107)

# Read the data
ebola <- read.csv("Ebola_outbreak_DRC2018_data.csv")
ebola$Date <- chron(as.character(ebola$Date), format=c(dates = "day mon year"))     

# Definition of the intervention function
exponential <- function(t,beta0,beta1,k,tau1) {
  if(t < tau1) {
    	beta <- beta0
	} else {
		beta <- beta0*beta1 + (beta0-beta0*beta1)*exp(-k*(t-tau1))
	}
	return(beta)
}
Exponential <- Vectorize(exponential, vectorize.args = 't')

# Definition of the SEIR model
SEIR <- function(t, x, parms) {
    with(as.list(c(parms,x)),{
		beta <- exponential(t,beta0,beta1,k,tau1)
        # Density-dependent transmission
		N <- S + E + I + R + D
		dS <- - beta/N*S*I
		dE <- beta/N*S*I - sigma*E
		dI <- sigma*E - gamma*I
		dR <- (1-f)*gamma*I
		dD <- f*gamma*I
		dC <- sigma*E
		der <- c(dS,dE,dI,dR,dD,dC)
		list(der)
	})
}

# Negative log-likelihood
nll <- function(beta0,beta1,k,f,tau0,tau1,sigma,gamma) {
	pars <- c(beta0=beta0,beta1=beta1,k=k,f=f,tau0=tau0,tau1=tau1,sigma=sigma,gamma=gamma)
	pars <- trans(pars)
	times <- c(data$times+pars["tau0"],max(data$times+pars["tau0"])+1)
	simulation <- as.data.frame(ode(init,times,SEIR,parms=pars))
	ll <- sum(dpois(data$cases,diff(simulation$C),log=TRUE))
	return(-ll)
}

# Parameter transformation
trans <- function(pars) {
	pars["beta0"] <- exp(pars["beta0"])
	pars["beta1"] <- plogis(pars["beta1"])
	pars["k"] <- exp(pars["k"])
	pars["f"] <- plogis(pars["f"])
	pars["tau0"] <- exp(pars["tau0"])
	if(is.na(pars["tau1"])) {
		pars["tau1"] <- pars["tau0"]
	}
	return(pars)
}
```

The Democratic Republic of Congo (DRC) currently experiences an outbreak of Ebola virus disease (EVD). According to [WHO situation reports](http://www.who.int/ebola/situation-reports/drc-2018/en/), the outbreak began in early April 2018 and remains localized to three health zones (Iboko, Bikoro and Wangata) in the north-west of DRC.

Real-time analyses of the transmission dynamics of this outbreak are important for anticipating further spread of EVD and investigating the effects of control interventions. The transmission potential of an infectious disease can be described by the basic reproduction number ($R_0$) that describes the average number of secondary infections generated by an infectious index case at the beginning of an outbreak. An outbreak can be brought under control once the net reproduction number $R_t$ drops below unity.

Based on the same methodology that we developed for analyzing the 2014 EVD outbreak in DRC [(Althaus, 2015)](https://peerj.com/articles/1418/), this document provides a real-time analysis of the current outbreak using the most recent data of reported cases from the WHO situation reports (Figure 1).

```{r plot_data, echo=FALSE, fig.height=5, fig.width=10}
weeks <- seq.dates(from = chron("05 Apr 2018", format = c(dates = "day mon year")), by = "week", length = ceiling(length(ebola$Cases)/7)+1)
barplot(ebola$Cases,xlab="Date of onset of symptoms",ylab="Number of cases",axes=FALSE)
axis(1,seq(from=0.7,by=7*1.2,length.out=ceiling(length(ebola$Cases)/7)+1),weeks)
axis(2)
```

**Figure 1. Confirmed and probable EVD cases in the Democratic Republic of Congo (DRC)** The [WHO situation report](http://apps.who.int/iris/bitstream/handle/10665/272825/SITREP_EVD_DRC_20180612-eng.pdf?ua=1) from 12 June 2018 reported 52 confirmed and probable EVD cases until 10 June 2018.

## Outbreak analysis and projection

```{r fit, echo=FALSE, cache=TRUE}
# Prepare the data and set the initial values
data <- na.omit(ebola[c("Date","Cases")])
names(data) <- c("times","cases")
begin <- chron("05 Apr 2018", format=c(dates = "day mon year")) 
data$times <- data$times - data$times[1]
data$cases[1] <- 0
N <- 1e6
init <- c(S = N - 1, E = 0, I = 1, R = 0, D = 0, C = 0)

# Fit the model to the data
fixed <- c(beta1 = qlogis(0), sigma = 1/9.312799, gamma = 1/7.411374, tau0 = -Inf, f = qlogis(0.5))
free <- c(beta0 = log(2/7.411374), k = log(0.1), tau1 = 30)
fit <- mle2(nll,start=as.list(free),fixed=as.list(fixed),method="Nelder-Mead",control=list(maxit=1e3))

# Parameter sampling (for CIs and prediction interval) 
n_sim <- 1e4
m <- coef(fit,exclude.fixed=TRUE)
sigma <- vcov(fit)
sim_coef <- data.frame(rmvnorm(n_sim, mean = m, sigma = sigma))

# Timepoints
pars <- trans(coef(fit))
end <- chron("12 Jul 2018", format=c(dates = "day mon year"))
weeks <- seq.dates(from = chron("05 Apr 2018", format = c(dates = "day mon year")), to = end, by = 7)
timepoints <- seq(0,end-begin,1)

# Simulations
last_dp <- length(data$cases)
mod_sim_Inc <- matrix(data=NA,nrow=length(timepoints),ncol=n_sim)
obs_sim_Inc <- matrix(data=NA,nrow=length(timepoints),ncol=n_sim)
mod_sim_Cum <- matrix(data=NA,nrow=length(timepoints)-last_dp,ncol=n_sim)
obs_sim_Cum <- matrix(data=NA,nrow=length(timepoints)-last_dp,ncol=n_sim)
for(i in 1:n_sim) {
    pars <- trans(c(unlist(sim_coef[i,]),fixed))
    simulation <- as.data.frame(ode(init,timepoints,SEIR,parms=pars))
    mod_Inc <- c(0,diff(simulation$C))
    mod_sim_Inc[,i] <- mod_Inc
    obs_Inc <- rpois(length(timepoints), lambda = mod_Inc)
    obs_sim_Inc[,i] <- obs_Inc
    mod_sim_Cum[,i] <- cumsum(mod_Inc[-(1:last_dp)]) + sum(data$cases)
    obs_sim_Cum[,i] <- cumsum(obs_Inc[-(1:last_dp)]) + sum(data$cases)
}
```

```{r plot_dynamics, echo=FALSE, fig.height=10, fig.width=10}
par(mfrow=c(2,1))
# Plot the daily incidence of new cases
mod_sim <- mod_sim_Inc
obs_sim <- obs_sim_Inc
x_grid <- timepoints
mod_q <- apply(mod_sim, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
obs_q <- apply(obs_sim, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
simulation <- as.data.frame(ode(init,x_grid,SEIR,parms=trans(coef(fit))))
plot(x_grid, mod_q[2,], ylim = c(0,7), ty="l", lty=2, lwd = 1, col = "red", xlab=NA, ylab="Daily incidence of new cases",frame=FALSE,axes=FALSE)
axis(1, weeks - begin, weeks)
axis(2)
polygon(x = c(x_grid, rev(x_grid)), y = c(obs_q[1,], rev(obs_q[2,])), col = rgb(1, 0, 0, alpha=0.2), border = NA)
lines(x_grid, mod_q[1,], lty = 2, lwd = 1, col = "red")
lines(x_grid, c(0,diff(simulation$C)), lwd = 1, col = "red")
points(data$times,data$cases,col="red")

# Plot the cumulative number of new cases
mod_sim <- mod_sim_Cum
obs_sim <- obs_sim_Cum
x_grid <- timepoints
mod_q <- apply(mod_sim, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
obs_q <- apply(obs_sim, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
simulation <- as.data.frame(ode(init,x_grid,SEIR,parms=trans(coef(fit))))
plot(x_grid[-(1:last_dp)], 1+mod_q[2,], xlim=range(x_grid), ylim = c(0,70), ty="l", lty=2, lwd = 1, col = "red", xlab=NA, ylab="Cumulative number of cases",frame=FALSE,axes=FALSE)
axis(1, weeks - begin, weeks)
axis(2)
polygon(x = c(x_grid[-(1:last_dp)], rev(x_grid[-(1:last_dp)])), y = c(1+obs_q[1,], rev(1+obs_q[2,])), col = rgb(1, 0, 0, alpha=0.2), border = NA)
lines(x_grid[-(1:last_dp)], 1+mod_q[1,], lty = 2, lwd = 1, col = "red")
lines(x_grid[-(1:last_dp)], 1+sum(data$cases)+simulation$C[-(1:last_dp)]-simulation$C[last_dp], lwd = 1, col = "red")
points(data$times,1+cumsum(data$cases),col="red")
```

**Figure 2. Dynamics of EVD outbreak in the Democratic Republic of Congo (DRC).** Model fits of daily incidence (upper panel) and projections for cumulative numbers (lower panel) of cases are shown together with reported data (circles). The best-fit model (solid lines) is given together with the 95% confidence intervals (dashed lines). The shaded areas correspond to the 95% prediction intervals.

```{r reproduction, echo=FALSE, cache=TRUE}
# Calculate the effective reproduction number vs. time (including approximate pointwise CI)
beta <- function(t, pars) Exponential(t, pars[["beta0"]],pars[["beta1"]],pars[["k"]],pars[["tau1"]])

pars <- trans(coef(fit))
timepoints <- seq(0,end-begin,0.1)
Re <- beta(timepoints, pars)/pars[["gamma"]]    
Re_sim <- matrix(data=NA,nrow=length(timepoints),ncol=n_sim)
for(i in 1:n_sim) {
  pars_i <- trans(c(unlist(sim_coef[i,]), fixed))
  Re_sim[,i] <- beta(timepoints, pars_i)/pars_i[["gamma"]]
}
Re_q <- apply(Re_sim, MAR = 1, FUN = quantile, probs = c(0.025, 0.975))
```

```{r plot_reproduction, echo=FALSE, fig.height=5, fig.width=10}
# Plot the effective reproduction number
plot(timepoints,Re,type="n",col="blue",ylim=c(0,4),xlab=NA,ylab=bquote("Net reproduction number " ~ italic("R"["t"])),axes=FALSE,frame=FALSE)
polygon(x = c(timepoints, rev(timepoints)), y = c(Re_q[1,], rev(Re_q[2,])), col = rgb(0, 0, 1, alpha=0.2), border = NA)
lines(timepoints, Re, col = "blue")
lines(timepoints, Re_q[1,], col = 'blue', lty = 2)
lines(timepoints, Re_q[2,], col = 'blue', lty = 2)
axis(1, weeks - begin, weeks)
axis(2, 0:5)
abline(h=1,lty=3)
points(28,Re[280],pch=19)
points(33,Re[330],pch=19)
points(46,Re[460],pch=19)
text(28,Re[280],"Initial report of the outbreak (3 May)",adj=c(-0.03,-0.1))
text(33,Re[330],"Official declaration of the outbreak (8 May)",adj=c(-0.03,-0.1))
text(46,Re[460],"Launch of vaccination exercise (21 May)",adj=c(-0.03,-0.1))
```

**Figure 3. Net reproduction number $R_t$ during the Ebola virus disease (EVD) outbreak in the Democratic Republic of Congo (DRC).** The maximum likelihood estimates of the net reproduction number $R_t$ (solid line) are shown together with the 95% confidence intervals (dashed lines). The dotted line denotes the critical threshold $R_t = 1$.
